Male-biased sexual selection, but not sexual dichromatism, predicts speciation in birds

### 2020-05-30 Manuscript_Publication.Rmd
Updated supp material

---
title: Male-biased sexual selection, but not sexual dichromatism, predicts speciation
  in birds
author:
bibliography: References/SS_Extinction.bib
output:
  word_document: default
  pdf_document:
    keep_tex: yes
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[L]{\textsl{\leftmark}}
  \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{}
  \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{}
  \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} \usepackage[hang,
  flushmargin]{footmisc} \usepackage{hyperref}\hypersetup{draft,colorlinks=false, hyperfootnotes=false,
  linkcolor=blue, filecolor=blue, urlcolor=blue} \usepackage{xcolor} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}}
  \raggedbottom
csl: References/evolution.csl
subparagraph: yes
abstract: Sexual selection is thought to shape phylogenetic diversity by affecting speciation or extinction rates. However, the net effect of sexual selection on diversification is hard to predict _a priori_, because many of the hypothesised effects on speciation or extinction have opposing signs and uncertain magnitudes. Here, we test whether the strength of sexual selection predicts variation in speciation and extinction rates across passerine birds (up to 5,812 species, covering most genera). We use two measures of sexual selection (sexual dichromatism and a multivariate measure of male-biased sexual selection) and adjust for range size and environmental variability. Male-biased sexual selection, but not sexual dichromatism, predicted speciation rates ($\lambda_{DR}$) in passerines. This relationship was independent of range size or environmental variability, though species with smaller ranges had higher speciation rates. There was no association between sexual selection and extinction rate ($\mu_{BAMM}$). Our findings suggest that sexual dichromatism is a poor proxy for sexual selection at least a broad-scale and support the view that sexual selection has shaped diversification in songbirds.
---
<!-- \maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Parkville, VIC, 3052, Australia} -->

>>>>

---
title: Male-biased sexual selection, but not sexual dichromatism, predicts speciation
  in birds
author:
bibliography: References/SS_Extinction.bib
output:
  pdf_document:
    keep_tex: yes
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[L]{\textsl{\leftmark}}
  \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{}
  \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{}
  \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} \usepackage[hang,
  flushmargin]{footmisc} \usepackage{hyperref}\hypersetup{draft,colorlinks=false, hyperfootnotes=false,
  linkcolor=blue, filecolor=blue, urlcolor=blue} \usepackage{xcolor} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}}
  \raggedbottom
csl: References/evolution.csl
subparagraph: yes
abstract: Sexual selection is thought to shape phylogenetic diversity by affecting speciation or extinction rates. However, the net effect of sexual selection on diversification is hard to predict _a priori_, because many of the hypothesised effects on speciation or extinction have opposing signs and uncertain magnitudes. Here, we test whether the strength of sexual selection predicts variation in speciation and extinction rates across passerine birds (up to 5,812 species, covering most genera). We use two measures of sexual selection (sexual dichromatism and a multivariate measure of male-biased sexual selection) and adjust for range size and environmental variability. Male-biased sexual selection, but not sexual dichromatism, predicted speciation rates ($\lambda_{DR}$) in passerines. This relationship was independent of range size or environmental variability, though species with smaller ranges had higher speciation rates. There was no association between sexual selection and extinction rate ($\mu_{BAMM}$). Our findings suggest that sexual dichromatism is a poor proxy for sexual selection at least a broad-scale and support the view that sexual selection has shaped diversification in songbirds.
---
<!-- \maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Parkville, VIC, 3052, Australia} -->

### 2020-05-30 Sexual Selection and Speciation.Rmd
Updated supp material

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(pander)
library(knitr)
library(tidyr)
library(mgcv)
library(diversitree)
library(repmis)
library(lme4)
library(ape)
library(geiger)
library(grid)
library(svglite)
library(gridExtra)
library(nlme)
library(phytools)
library(brms)
library(ggridges)
library(caper)
library(purrr)
library(reshape2)
library(ggExtra)
library(grid)
library(dplyr)
library(BAMMtools)
library(gridExtra)
library(car)
library(coda)
library(MuMIn)
library(parallel)
library(HDInterval)
library(phangorn)
library(kableExtra)
library(RColorBrewer)
library(tibble)
library(stargazer)
library(ggraph)
library(ggiraphExtra)
library(stringr)
library(Cairo)
library(gdtools)
library(data.table)
library(bindrcpp)
#devtools::install_github("Ax3man/phylopath")
#devtools::install_github("thomasp85/patchwork")
#devtools::install_github('VPetukhov/ggrastr')
library(ggrastr)
library(patchwork)
library(phylopath)
if(!requireNamespace("BiocManager", quietly = TRUE)){
    install.packages("BiocManager")
  }
#BiocManager::install(c("EBImage", "ggtree"))
library("EBImage") # for images
library("ggtree")
library("ggnewscale")
source("bamm_extinction/functions/check_and_fix_ultrametric.R")
source("functions/essim.R")
source("functions/diversification_rate_calculation.R")
source("functions/Vdodge_function.R")
source('functions/symmetrise_scale.R')
source('functions/Correlation_matrix.R')
source('functions/gheatmap2.R') #Gets rid of spaces between tiles
knitr::opts_chunk$set(cache=TRUE)
```

@@

```{r trycatch all models, eval = F}

@@

SS.subset %>% ggplot(aes(x = SDi, y = Sexual_selection_ppca))+
  geom_point()+
  geom_smooth(method = "lm")+
  theme_minimal()
```
**Figure S12:** The relationship between absolute sexual dichromatism and male-baised sexual selection. Although positively correlated, the distribution is very noisy with some species having high sexual dichromatism and low male-biased sexual selection and vice versa.

@@

#Set rownames to match tree
rownames(SS.subset) <- SS.subset$TipLabel

@@

```{r random calcs two, eval = FALSE, echo = F}
#Run subset model with SDi
#Run model for DR
Subset.MCC.DR.SDi <- gls(MCC.DR ~ SDi
                         + log(range.size.m2)
                         + bioclim4 #Seasonal variation
                         + residuals.PC1 #Spatial variation
                         + PC1.LIG #Long-term climate variation
                         + NPP,
                correlation = corPagel(1, phy = pruned.MCC.Subset.tree, fixed = FALSE), 
                data = SS.subset, 
                method = "REML")

@@

>>>>

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(pander)
library(knitr)
library(tidyr)
library(mgcv)
library(diversitree)
library(repmis)
library(lme4)
library(ape)
library(geiger)
library(grid)
library(svglite)
library(gridExtra)
library(nlme)
library(phytools)
library(brms)
library(ggridges)
library(caper)
library(purrr)
library(reshape2)
library(ggExtra)
library(grid)
library(dplyr)
library(BAMMtools)
library(gridExtra)
library(car)
library(coda)
library(MuMIn)
library(parallel)
library(HDInterval)
library(phangorn)
library(kableExtra)
library(RColorBrewer)
library(tibble)
library(stargazer)
library(ggraph)
library(ggiraphExtra)
library(stringr)
library(Cairo)
library(gdtools)
library(data.table)
library(bindrcpp)
#devtools::install_github("Ax3man/phylopath")
#devtools::install_github("thomasp85/patchwork")
#devtools::install_github('VPetukhov/ggrastr')
library(ggrastr)
library(patchwork)
library(phylopath)
library(sjPlot)
library(sjmisc)
if(!requireNamespace("BiocManager", quietly = TRUE)){
    install.packages("BiocManager")
  }
#BiocManager::install(c("EBImage", "ggtree"))
library("EBImage") # for images
library("ggtree")
library("ggnewscale")
source("bamm_extinction/functions/check_and_fix_ultrametric.R")
source("functions/essim.R")
source("functions/diversification_rate_calculation.R")
source("functions/Vdodge_function.R")
source('functions/symmetrise_scale.R')
source('functions/Correlation_matrix.R')
source('functions/gheatmap2.R') #Gets rid of spaces between tiles
knitr::opts_chunk$set(cache=TRUE)
```

@@

```{r trycatch all models, eval = F}
#Read in DR models and extract estimates:

@@

SS.subset %>% ggplot(aes(x = SDi, y = Sexual_selection_ppca))+
  geom_point()+
  geom_smooth(method = "lm")+
  theme_minimal()

#Set rownames to match tree
rownames(SS.subset) <- SS.subset$TipLabel
```
**Figure S12:** The relationship between absolute sexual dichromatism and male-baised sexual selection. Although positively correlated, the distribution is very noisy with some species having high sexual dichromatism and low male-biased sexual selection and vice versa.

@@

@@

## Testing the sexual dichromatism dataset using the same data points as the sexual selection dataset

In our analysis we conducted tested for the effect of sexual dichromatism on speciation using a dataset with a sample size of 5,812 passerine bird species. However, when testing for the effects of sexual selection (using a principal component score) we only utilised 2,465 species. This discrepancy is because data for the principal components analysis is incomplete for many species. Here we conduct an analysis on the effects of sexual dichromatism on speciation using the same model structure as in our primary analysis, however using the same data/species as we used in the analysis of the effect of sexual selection (*n = 2,465*). We conduct the analyses on MCC trees for $\lambda_{DR}$, $\lambda_{ND}$ and $\lambda_{BAMM}$ and compare the results to the analyses conducted on 5,812 passerine bird species.

```{r subset SDi, eval = TRUE, warning=FALSE, message=FALSE, fig.height=12, fig.width=8}
if(!file.exists("data/sample_size_comparison.rds")){
# Read in tree 
pruned.MCC.Subset.tree <- readRDS('data/pruned.MCC.Subset.tree.rds')

#Run subset model with SDi
#Run model for DR
Subset_SDi_Models <- list()

Subset_SDi_Models[["DR"]] <- gls(MCC.DR ~ SDi
                               + log(range.size.m2)
                               + bioclim4 #Seasonal variation
                               + residuals.PC1 #Spatial variation
                               + PC1.LIG #Long-term climate variation
                               + NPP,
                      correlation = corPagel(1, phy = pruned.MCC.Subset.tree, fixed = F), #Subset Tree 
                      data = SS.subset, 
                      method = "REML")

#Run model for ND
Subset_SDi_Models[["ND"]] <- gls(MCC.ND ~ SDi
                               + log(range.size.m2)
                               + bioclim4 #Seasonal variation
                               + residuals.PC1 #Spatial variation
                               + PC1.LIG #Long-term climate variation
                               + NPP,
                         correlation = corPagel(1, phy = pruned.MCC.Subset.tree, fixed = F), #Subset Tree 
                         data = SS.subset, 
                         method = "REML")

#Run model for ND
Subset_SDi_Models[["BAMM"]] <- gls(mean.lambda ~ SDi
                                 + log(range.size.m2)
                                 + bioclim4 #Seasonal variation
                                 + residuals.PC1 #Spatial variation
                                 + PC1.LIG #Long-term climate variation
                                 + NPP,
                           weights = ~ sqrt(var.lambda),
                           correlation = corBrownian(phy = pruned.MCC.Subset.tree), #Subset Tree
                           data = SS.subset, 
                           method = "REML")

# Main analysis models 
Main_SDi_Models <- list()

Main_SDi_Models[["DR"]] <- readRDS('data/MCC.DR.top.rds')
Main_SDi_Models[["ND"]] <- readRDS('data/MCC.ND.top.rds')
Main_SDi_Models[["BAMM"]] <- readRDS('data/MCC.Lambda.top.rds')

sample_size_comparison <- list(Subset_SDi_Models = Subset_SDi_Models, Main_SDi_Models = Main_SDi_Models)
saveRDS(sample_size_comparison, "data/sample_size_comparison.rds")
} else {
  sample_size_comparison  <- readRDS("data/sample_size_comparison.rds")
}

plot_mods <- function(model1, model2) {

}
sjplot_custom <- function(sub, full, type, min, max) {
  theme_set(theme_sjplot())
plot <-  sjPlot::plot_models(sub, 
                      full, 
                      show.p = FALSE, 
                      transform = NULL, 
                      std.est = "std2", 
                      m.labels = c("Subset data (n = 2,465)", "Full data (n = 5,812)"), 
                      legend.title = "Data Points Used",
                      title = paste("Comparing", type, "models with subset data")) + 
    ylim(min, max) +
    theme_sjplot2() + 
  scale_color_sjplot("simply")
  return(plot)
}

grid.arrange(
sjplot_custom(sample_size_comparison[["Subset_SDi_Models"]][["DR"]], 
              sample_size_comparison[["Main_SDi_Models"]][["DR"]], 
              type = "DR", min = -0.025, max = 0.025),
sjplot_custom(sample_size_comparison[["Subset_SDi_Models"]][["ND"]], 
              sample_size_comparison[["Main_SDi_Models"]][["ND"]], 
              type = "ND", min = -0.00075, max = 0.00075),
sjplot_custom(sample_size_comparison[["Subset_SDi_Models"]][["BAMM"]], 
              sample_size_comparison[["Main_SDi_Models"]][["BAMM"]], 
              type = "Lambda-BAMM", min = -0.002, max = 0.002))

@@

**Figure S17:** A comparison of full and subset data was used in models testing for the effect of sexual dichromatism on speciation rate ($\lambda_{DR}$, $\lambda_{ND}$ and $\lambda_{BAMM}$). While there is variation between model estimates due to the different data points used, these differences do not lead us to make alternive conclusions. In all three cases the partial regression coefficient for sexual dichromatism (SDi) has 95 % confidence intervals that overlap zero, thus we have no reason to suspect that the effect that this study found of sexual selection on speciation is due to sampling bias.
